Augmentation of a Model Historical Harpsichord Mechanism for Haptic Enabled Interaction

### 2025-01-20 src/article.tex
Updates from Overleaf

The Tagliavini Collection in Bologna \cite{Tagliavini2007}, renowned for its historical keyboard instruments, exemplifies this dilemma. To address it, this project introduces a haptic keyboard, a digital interface designed to bridge the gap between preservation and accessibility. The project builds on Andrew McPherson’s earlier work on "Portable Measurement and Mapping of Continuous Piano Gesture" and aligns with principles outlined in The O in NIME, which emphasises the importance of extending the lifecycle of existing NIMEs (New Interfaces for Musical Expression). By reusing, updating, and complementing previous NIMEs, the project fosters sustainability and ensures their relevance beyond their original context.

The haptic keyboard combines high-definition sound sampling, touch-sensitive technology, and ergonomic design, recreating the sound and tactile qualities of the Tagliavini instruments without compromising their physical integrity.

@@

This work inherits the same philosophy as the Tromba Moderna project. By augmenting a replica of a historical harpsichord keyboard using minimally invasive electronics, the project intends to offer a tool enhancing the fruition of the Tagliavini collection whilst retaining a form of continuity with historical musical instrument-building traditions.

Beyond enhancing the visitor experience, the interface serves as a research tool to explore the unique characteristics of the harpsichord and its impact on performance.

@@

The model was designed to have two internal chambers a front and rear.
The front and back that was xxx by yyy by 370mm between the faceplate
and the jacks. The TONGUE THAT GOES INTO SLOT divides the chamber though
a small gap between the under side of the soundboard and the key guides.
The key guide needs to be deep enough to allow for the full movement of
the key and stop the key becoming unseated. The chamber then extends to
a support wall which divides the front and rear.

@@

The rear chamber was designed to house components for audio processing,
which is discussed further in \textbackslash section

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{images/49-key-bottom.jpeg} 
  \caption{Underside of the full model keyboard showing two chambers. Front chamber at the top and rear chamber at the bottom.} 
  \Description{} 
  \label{fig:49-key-bottom}
\end{figure}

@@

For the final interface a 49-key (4-octave) was constructed by luthier
Roberto Livi. The keyboard is in the same style† as the 3-key model with
to jacks per key.

@@

Timescale: from initial discussion to delivery of interface to the NEMUS
lab was 8-months. Active assembly of the instrument was 3 months

@@

A set of criteria was drawn to which the final sensor system would have
to satisfy. This criteria was used as a means of evaluating the
feasability throughout the prototyping phase.

@@

\section{Hardware Design}\label{hardware-design}

The hardware was designed with the xx in mind.

@@

\subsection{Failed sensors}\label{failed-sensors}

A number of sensors were tried before the final system was decided upon.
An inertial measurement unit (IMU) was mounted to a single key which
contained a 6-degrees-of-freedom accelerometer and gyroscope. There was
a consistent set of conditions to trigger sound, but this was not easily
differentiable when other keys were played. The fitting time and cost of
a single IMU could b reduced when purchased in bulk but was far greater
than other systems considered.

A reed sensor and a hall sensor were mounted and magnets embedded within
the jacks. Adjusting the height of the sensor provided a reliable means
of detecting a threshold, in particular the hall sensors which could be
partially calibrated physically and also in software. The time taken to
calibrate a single key was time consuming. Ideally each sensor would be
placed on a thread for adjustment, but given the constraints on the
internal space it was decided any adjustment of heights would likely
mean an unacceptable length of time to install and calibrate. Embedding
magnets into the jacks was also did not satisfy the criteria of being
scalable and non-invasive.

Light dependant resistor mounted to the jack rail. as jacks were pressed
less light would be let through. Usable data but very dependant on
lighting conditions. Modifying the jack also did not satisfy the
non-intrusive criteria and the placement outside the internal chamber of
the model meant hiding electronics would be more difficult.

Finally, force sensitive resistors placed under the jack of each key was
considered but was never implemented. The sensors were to be placed
under each jack either above or below the padding on which the bottom of
the jack rests.

@@

Design of electronics carried out in tandem with construction of full
model. Since the Electronics were designed at a distanced to the final
full model some flexbility was required in the design. Design decisions
that were not as optimal but would allow for change in wiring and
fitting should there be any issues during installation.

Electronics designed so that they would not require any equipment
outside what can commonly be found in a maker space at a univeristy.

The equipment required including soldering suitable for throughole
technology (THT) components and 3D printing.

Laser cutting and CNC routing fascilities were used during the research
for the project but not required for the final implementation.

@@

NEMUS lab under construction during the first section of the project.
Facilities available at the maker spaces of Bologna (Alma Labor) and
Edinburgh (uCreate) used for what was deemed reasonable assumption in
terms of available equipment.

\begin{itemize}
\item
  Over arching themes

  \begin{itemize}
  \item
    simplicity of design

    \begin{itemize}
    \item
      Limited resources
    \item
      easily reproducible
    \end{itemize}
  \item
    availability and lead times
  \item
    iteration and design from distance
  \end{itemize}
\end{itemize}

Following from the work in \cite{McPherson2013, McPherson2019} we tested
a system using the Fairchild QRE1113 \textbackslash datasheetfigue. The
QRE1113 is a combination infrared LED and phototransistor sensitive to
IR light. The phototransistor of QRE1113 provides data on how much light
is present and in particular how much light is being relfected from a
surface. Since refelected light will be proportional to the distance of
a surfcae it is a good, close proximity, distance sensor. Distance is
the way in which the instruments in \cite{McPherson2013, McPherson2019}
function as well the Moog Piano Bar from which it too inspiration.

The limitation for taking a distance approach is that there is not one
but 2 jacks per key. For each jack to be measured indepeedntly the data
would need tp com fropm the jacks themselves.

\begin{itemize}
\item
  The triggering would need to be bespoke for every key as fluxuations
  in the dimensions of every piece mean that the threshold would be
  different.

  \begin{itemize}
  \item
    Moving a problem into additional time in the calibtraryoin setup
    phase.
  \end{itemize}
\item
  Remebering the constraint that system isshould noyt fix problem, there
  would be no easy way to tell if a jack had nbot reseated.

  \begin{itemize}
  \item
    In playing a harpihchord when the jack does not go to its rest
    position, not note would play. This fault should be carried over
    intp the final instrument.
  \end{itemize}
\end{itemize}

Another common application of the QRE1113 is within line following
robots, where an array of QRE1113s ar used in conjunction with black
tape along which some motorised system can following

An initial sticker design was used that contained only the gradient. It
was found that the jacks wer sticking in their slots andthe edges on the
stickers easily peeled around the printing The sticker was redesigned to
be the full length of the jack which avoided the problem of peeling. The
larger are also mean the stickers adhered betterto the jack Some
trimming was still required of each sticker to avoid them adding extraa
resistence to the jack's movement.

During testing it was found that their was a considerable amount
cross-talk bwteen adjacent sensors. This meant the reading from key
would be affected by the movement of the key next to it. The cross-talk
was assumed to be as a result of light reflecting from other jacks.
Given that IR is used this was difficult to confirm by sight. A set of
baffles was designed for the pcb that would wrap round the jack and stop
light from other jacks or sensors. Baffles were printed on BAMBU X1 3D
printers using PLA plastic. A dark pigmented PLA was used to avoid IR
light simply being diffused. After the baffles were fitted the
cross-talk was entirely eliminated.

\subsection{Arduino}\label{arduino}

\subsubsection{Nano BLE}\label{nano-ble}

The Nano BLE specifically was used in the initial stages for the
integrated IMU.

When it was finalised that the prokect would proceed using the QRE1113 a
number of microcontroller units (MCU) were condsidered.

A rough performance test was later carried out using an STM32
NUCLEO-L031K6, a Nano ESP32 and the BLE. The Nano BLE was found to be
more performative. To execute a single sensor read cycle of the firmware
the STM32 took 40ms, and both the Nano ESP32 and Nano BLE required 11ms
on average. The Nano ESP32 had high variability so it was decided to
continue with the Nano BLE. Further discussion of using ESP32 continues
in \textbackslash section\{Going forward\}

\subsection{EEPROM / NVRAM / FRAM}\label{eeprom-nvram-fram}

There would need to be the ability to store values on non-volatile
memory so that the electroincs could be powered-down without losing
data.

\emph{It is worth noting that the FRAM model chosen stopped being
available during the project, meaning any future iteration would need to
adjust for another board in firmware}

The firmware was structure in a way that the interface could be changed
so long as the code required for any other board would comply with
function arguments and outputs. Worth noting the fragility and
volatility when designing this kind of project and how easily it is for
the project to become broken.

SD card would provide an easier way to interact with calibration data.
The posiiton of the controller board did not permit easy access.
reliability and speed was preffered over storage capacity.

\subsection{RGB LEDs}\label{rgb-leds}

The LEDs also allowed for easy identification for keys that were outside
of expected values and when a key was triggered. This allowed for visual
identification of malfunctioning or uncalibrated keys.

The LEDs are on the same PCB as the sensors and are hidden when the
nameboard is in place.2024

Would sugggest that this is a necessity of this kind to provide a means
to identify and diagnose LEDs with state.

TLC5940 or other PWM driver such as in \cite{McPherson2013} added
complexity to assembling, but were avoid in favour of Addressable LEDs
with intergrated driver. The intergrated LEDs share the same data line
which has the limitation where a fualty LED will stop all following LEDs
from working. An individual PWM driver per PCB would allow avoid
situations where failutre cascades, but would add complexity in assembly
as the parts are typically SMD. Also, given the LEDs are fixed to the
board and could not be repaired in situ, the solution would be to
replace the PCB regardless of which approach was taken. Given the LEDs
were only used in debugging and calibration and otehrwise would not be
visible, a simpler design approach was taken.

\subsection{Multiplexing}\label{multiplexing}

CD4051BE Texas Instruments

problem of scale

For the 3-key model, individual jacks could be connected to one of the 8
available ADC pins.

The BLE Nano has 8 ADC pins connected to one ADC channel via a
multiplexer

Multiplexer added to expand to the

Transistor / Op-amp circuit could have been used similar to the one in
\cite{McPherson2013}.

Benefits were keeping with THT based designed for easy assembly.

An 8-channel multiplexer was chosen as the model has 49-keys and a
division of 7 PCBs each containing 7 sensors

The multiplexer reduced 7 signals to a single pin. Three pins are
required pins for addressing the multiplexers and all PCBs are addressed
simultaneously.

\subsection{PCB design}\label{pcb-design}

EAGLE PCBs designed using EAGLE. Worth noting that AutoCAD has
deprecated EAGLE and the programme will no longer be supported from
2026 Project files will need to be transitioned to another standard,
likely the KiCAD programme.

trade off between limiting number of boards required and allowing
compensation for variation in pitch.

Smaller PCBs means that spacing could be adjusted to accomodate the
changes in pitch between jacks. It also allowed for a modular approach
where any PCB could swapped for minimal cost.

\subsubsection{Sensor board}\label{sensor-board}

\begin{itemize}
\item
  x7 QRE1113
\item
  x7 100R
\item
  x7 10kR
\item
  x1 CD4051BE Multipexer
\item
  jack / sensor pitch by board
\item
  7 boards adjust for pitch/spacing error
\item
  modular, replaceable
\item
  orientation problems

  \begin{itemize}
  \item
    legs break easily
  \item
    end mount like A.P.Mc. time consuming
  \item
    problem

    \begin{itemize}
    \item
      milling out keys
    \end{itemize}
  \end{itemize}
\item
  designed from distance
\item
  striped lines across pcbs
\item
  problems

  \begin{itemize}
  \item
    additional terminals uneeded resulted in lots of additional work
  \item
    long cabling increase risk of breaking
  \end{itemize}
\item
  solutions

  \begin{itemize}
  \item
    striping all ADC lines with FFC
  \item
    solder pads to select channel
  \end{itemize}
\end{itemize}

\subsubsection{controller board}\label{controller-board}

@@

The 2.54mm pitch headers had two major drawbacks. First, the orientation
of the cables could be reversed, which in the best case scenario would
simply mean the system did not work correctly and in the worst case it
could potentially damaged some components.

@@

A locator was adapted from a creative commons STL model
\footnote{\url{https://www.thingiverse.com/thing:1646016}} that allowed a
generic crimping tool to be adapted for use with JST-PH crimps. 
The locator was printed using a Bambu X1 3D printer. 
A second alterntaive was assembling cables from other sets that were at hand.

@@

Firmare checks components are connected then reads the FRAM to see if
threshold data is already present. The rotary encoder and it's momentary
switch are polled for any change in state. A key is selected with the
rotary encoder while the momentary switch is used for changing between
states for key selection and threshold adjustment A double click of the
rotary encoder writes the current threshold values to FRAM. The sensors
for the current multiplexer channels are polled. If any sensor reads
above the threshold when it was previously below a MIDI On message is
sent. For the reverse case where the value was below the threshold when
it was previously above, a MIDI Note Off message ois sent.

@@

The final assembly took place at the NEMUS lab at the university of
Bologna over the final week in October 2024 for delivery at San
Colombano for exhibition in the first week of November in 2024.

@@

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{images/side_w-o_sensors.png} 
  \caption{} 
  \Description{} 
  \label{fig:3key}
\end{figure}

@@

The PCBs shared power and data lines for components with the exception
of the output from the multiplexer. Ribbon cable was cut to size to
allow for some room to readjust the distance of the sensors from the
jacks. The data signals from each PCB were connected using a separate
cable loom. Soldering cables and inspecting each join required another
day work. This did allow for readjustments to the positon of the
controller board Measurements of the rear chamber did not match with the
MacMini that was to be used a different layout was decided for fitting
the controller board/ Cutting cables to size allowed for a great deal of
felxibility but at the cost of time. After the first install it was
clearer what constraints were in place for fittig and an alternative
wiring system using flat flex cable was implement (discussed in
\textbackslash section)

@@

As the PCBs were fitted the jacks had the gradient stickers applied to
them. The jacks had been coated with a varnish on the short side to
provide a better surface for the stikcers to adhere to. After the
stickers were applied excess material was trimmed from each jack. The
jack were then check that they moved freely in their slot.

After fitting the PCBs preliminary calibration was carried out. Before
the keys were reffitted the jacks wer put back in place. The data from
each key was confirmed to be in a similar range. Any key showing
anomalous dat had it's jack removed and re-inspected. The PCB fitting
process and refitting required another day as faults were found in
wiring or misalignement of the stickers. After the sensors were
confirmed to be functioning the keys were put back.

The calibration workflow was with Dr.~Craig Webb who was unfamiliar with
the system. Adjustments were made to the firmware to to make it more
intuitive for the user. Addressable LEDs became vital as it allowed for
easy identification of the key being calibrated. They were also used to
identify any keys whose current reading was beyond what was expected.

@@

A process was implemented where the key was selected by playing, but
this caused problems in situations where the data was unusual due to
misalignement of the sensors

This would likely be a better system for when the keyboard was fully
operational

The thresholds were calibrated visually utilising the serial plotter of
the Arduino IDE and MIDI was confirmed using the open source MIDI
Monitor \footnote{https://github.com/krevis/MIDIApps}.

@@

Finer calibration carried out with the guidance of Catalina Vincens as
an expert harpsichord performer

@@

A limitation in the design is the variablility in positioning of tage,
especially when they need to be reapplied. A jig was designed for
applying tags, but could not be printed in time for installation. Even
with a jig, some tags had to be radjusted by hand regardless. Long term
this hasd impliccation on how to maintain the interface.

Taken a different approach to what surface is used. Preliminary test
were taken using Laser Cutting and CNC to fabricate Jack bodies. Laser
cutting results in a loss of material that has too much ariablity. The
edge o f the jack body will be charrred. The material also needs to be
reliable. on the advice of two luthiers, roberto livi and Jonathan Santa
Maria Bouquet at Edinburgh, Plywood and MDF have problems with
variations in humidity, tear-out and delamination.

Some tests were carried out using a notched jack body that was 3D
printed. The notch means that as the jack raises, the forward material
acts like a shutter. No reasons that this could not be CNC though a
variety of materials would be needed. Traditional material used is xx
though need to test of the material is naturally reflective enough or if
a coating would beed to be applied.

@@

A second interface designed more closely to the Trasuntino style
instrument commisioned by the NEMUS project with Roberto Livi. Using
lessons learned for the first iteration the second can make better
acommodations for internal electronics. Interface to be used for a
performer study and used for performance with a numerically simulated
sound syntheiss engine.

@@

Future novel designs for an interface that would utilise velocity and
aftertouch messages in software. This would be applicable for numericla
simulations of strings. A bespoke piec e of sowftare would also allow
for direct mapping of jack rows to registers. There is enough data to
implemenet functionality for velocity and aftertouch MIDI messages and
simply needs to be activated in the firmware. (Hyperinstruments
\cite{nime2024_20})

@@

Using the Espressif development framework ESP-IDF and follwong the same
logic as the original firmware source, the latency was reduced to around
10ms. ESP-IDF allows access to functionality of the ESP32 that is not
exposed in the Arduino libraries. It is possible that the 10ms latency
could be improved upon but there would be a cost in the form of
develoment time. Consideration would need to be made on how to port
existing code or whether to completely re-write the firmware.

@@

  \begin{itemize}
  \item
    debug if problem is power related Add power small power LEDs to the
    each PCB. It was not common for a board to not be functioning and
    thye result to be that the power weas not correctly connected.
    Especially when the ower is connected via somethgin such as an FFC
    cable where it is easy to plug-in the wrong way esepcially when
    fitting is to be carried out by someone unfmailiar wiyth the system.
    Power LEDs would quickly confirm whether the problem was power
    related. They would have to be small enough and low enough current
    draw to avoid glare or being mistaken with other indcator LEDs To
    continue using THT components, the choice would have to be moade
    carefully.
  \end{itemize}
\end{itemize}

>>>>

The Tagliavini Collection in Bologna \cite{Tagliavini2007}, renowned for its historical keyboard instruments, exemplifies this dilemma. To address it, this project introduces an augmented replica of a historical harpsichord keyboard. 
% The project builds on an existing design borrowed from \cite{McPherson2013} and aligns with principles outlined in \cite{Masu_NIME_2023}, emphasising the importance of extending the lifecycle of existing NIMEs. 
The project fosters ``sustainability of the results'', as identified in a previous NIME contribution by Masu \emph{et al.}  \cite{Masu_NIME_2023}, by updating and complementing the augmented piano keyboard design presented by McPherson \cite{McPherson2013} and ensuring the relevance of previous NIME designs beyond their original context.

@@

This work inherits the same philosophy as the Tromba Moderna project. By augmenting a replica of a historical harpsichord keyboard using minimally invasive electronics, the project intends to offer a tool enhancing the fruition of the Tagliavini collection whilst retaining a form of continuity with historical musical instrument-building traditions. Furthermore, the electronics design borrows ideas from a previous NIME by McPherson \cite{McPherson2013}. As such, the interface proposed here is not, strictly speaking, a \emph{new} musical interface, particularly in its haptic response designed to be as similar as possible to a real Italian harpsichord. However, this work complements and follows up on previous reflections regarding the \emph{sustainability of results} within the NIME community and identified as \emph{the O in NIME} \cite{Masu_NIME_2023}. The intended use of the keyboard through a meaningful bla bla is where the novelty of this work lies, rather than solely in its technological development.

Beyond enhancing the visitor experience, future design iterations will serve as a research tool to explore the unique characteristics of the harpsichord and its impact on performance through upcoming projects such as the ERC-funded Synergy grant Rem@ke \cite{remake1}.

@@

The model was designed to have two internal chambers a front and rear.
The front and back that was xxx by yyy by 370mm between the faceplate
and the jacks. The TONGUE THAT GOES INTO SLOT divides the chamber through
a small gap between the underside of the soundboard and the key guides.
The key guide needs to be deep enough to allow for the full movement of
the key and stop the key becoming unseated. The chamber then extends to
a support wall which divides the front and rear.

@@

% The rear chamber was designed to house components for audio processing,
% which is discussed further in \textbackslash section

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{images/49-key-bottom-} 
  \caption{Underside of the full model keyboard showing two chambers. Front chamber at the top and rear chamber at the bottom.} 
  \Description{} 
  \label{fig:49-key-bottom}
\end{figure}

@@

For the final interface a 49-key (4-octave) was constructed by luthier
\anon{Roberto Livi}. The keyboard is in the same style† as the 3-key model with
to jacks per key.

@@

Timescale: from initial discussion to delivery of interface to the \anon{NEMUS}
lab was 8-months. Active assembly of the instrument was 3 months

@@

A set of criteria was drawn to which the final sensor system would have
to satisfy. This criteria was used as a means of evaluating the
feasibility throughout the prototyping phase.

@@

% \section{Hardware Design}\label{hardware-design}

% The hardware was designed with the xx in mind.

@@

% \subsection{Failed sensors}\label{failed-sensors}

% A number of sensors were tried before the final system was decided upon.
% An inertial measurement unit (IMU) was mounted to a single key which
% contained a 6-degrees-of-freedom accelerometer and gyroscope. There was
% a consistent set of conditions to trigger sound, but this was not easily
% differentiable when other keys were played. The fitting time and cost of
% a single IMU could b reduced when purchased in bulk but was far greater
% than other systems considered.

% A reed sensor and a hall sensor were mounted and magnets embedded within
% the jacks. Adjusting the height of the sensor provided a reliable means
% of detecting a threshold, in particular the hall sensors which could be
% partially calibrated physically and also in software. The time taken to
% calibrate a single key was time consuming. Ideally each sensor would be
% placed on a thread for adjustment, but given the constraints on the
% internal space it was decided any adjustment of heights would likely
% mean an unacceptable length of time to install and calibrate. Embedding
% magnets into the jacks was also did not satisfy the criteria of being
% scalable and non-invasive.

% Light dependant resistor mounted to the jack rail. as jacks were pressed
% less light would be let through. Usable data but very dependant on
% lighting conditions. Modifying the jack also did not satisfy the
% non-intrusive criteria and the placement outside the internal chamber of
% the model meant hiding electronics would be more difficult.

% Finally, force sensitive resistors placed under the jack of each key was
% considered but was never implemented. The sensors were to be placed
% under each jack either above or below the padding on which the bottom of
% the jack rests.

@@

Design of electronics carried out in tandem with construction of full
model. Since the Electronics were designed at a distanced to the final
full model, some flexibility was required in the design. Design decisions
that were not as optimal but would allow for change in wiring and
fitting should there be any issues during installation.

Electronics designed so that they would not require any equipment
outside what can commonly be found in a maker space at a university.

The equipment required including soldering suitable for through-hole
technology (THT) components and 3D printing.

Laser cutting and CNC routing facilities were used during the research
for the project but not required for the final implementation.

@@

NEMUS lab under construction during the first section of the project.
Facilities available at the maker spaces of \anon{Bologna (Alma Labor)} and
\anon{Edinburgh (uCreate)} used for what was deemed reasonable assumption in
terms of available equipment.

\subsection{Controller Board}\label{controller-board}

@@

The 2.54mm pitch headers had two major drawbacks. First, the orientation
of the cables could be reversed, which in the best case scenario would
simply mean the system did not work correctly and in the worst case it
could potentially damage the components.

@@

A locator was adapted from a creative commons STL model
\footnote{\url{https://www.thingiverse.com/thing:1646016}} that allowed a
generic crimping tool to be adapted for use with JST-PH crimps. 
The locator was printed using a Bambu X1 3D printer. 
A second alternative was assembling cables from other sets that were at hand.

\subsubsection{Arduino Nano BLE}\label{arduino-nano-ble}

The Nano BLE was used in the initial stages for its integrated inertial measurment unit.

When it was finalised that the project would proceed using the QRE1113 a
number of microcontroller units (MCU) were considered.

A rough performance test was later carried out using an STM32
NUCLEO-L031K6, a Nano ESP32 and the BLE. The Nano BLE was found to be
more performative. To execute a single sensor read cycle of the firmware
the STM32 took 40ms, and both the Nano ESP32 and Nano BLE required 11ms
on average. The Nano ESP32 had high variability so it was decided to
continue with the Nano BLE.

% Further discussion of using ESP32 continues
% in \textbackslash section\{Going forward\}

\subsubsection{EEPROM / NVRAM / FRAM}\label{eeprom-nvram-fram}

There would need to be the ability to store values on non-volatile
memory so that the electronics could be powered-down without losing
data.

% \emph{It is worth noting that the FRAM model chosen stopped being
% available during the project, meaning any future iteration would need to
% adjust for another board in firmware}

% The firmware was structured in a way that the interface could be changed
% so long as the code required for any other board would comply with
% function arguments and outputs.

% Worth noting the fragility and
% volatility when designing this kind of project and how easily it is for
% the project to become broken.

SD card would provide an easier way to interact with calibration data.
The position of the controller board did not permit easy access.
reliability and speed was preferred over storage capacity.

\subsection{Sensor board}\label{sensor-board}

A number of sensors were tried before the final system was decided upon.
Sensors included, an 6-degrees-of-freedom inertial measurement unit,
reed and hall sensors paired with magnets, and light-dependant resistors mounted to the jack rail.
None of these sensors satisfied the sensor criteria described above.

Following from the work in \cite{McPherson2013, McPherson2019}, we tested
a system using the Fairchild QRE1113 \footnote{Fairchild, now ON Semiconductor, QRE1113 datasheet \url{https://www.onsemi.com/download/data-sheet/pdf/qre1113-d.pdf}}. The
QRE1113 is a combination infrared LED and phototransistor sensitive to
IR light. The phototransistor of QRE1113 provides data on how much light
is present and in particular how much light is being reflected from a
surface. Since reflected light will be proportional to the distance of
a surface it is a good, close-proximity, distance sensor. Distance is
the way in which the instruments in \cite{McPherson2013, McPherson2019}
function as well the Moog Piano Bar from which it took inspiration.

The limitation for taking a distance approach is that there is not one
but 2 jacks per key. For each jack to be measured independently the data
would need to come from the jacks themselves.

Another common application of the QRE1113 is within line following
robots, where an array of QRE1113s are used in conjunction with black
tape or reflective. A motorised system can then use this data to follow the line of tape and adjust its position.

An initial sticker design was used that contained only the gradient. It
was found that the jacks were sticking in their slots and the edges on the
stickers easily peeled around the printing The sticker was redesigned to
be the full length of the jack which avoided the problem of peeling. The
larger are also mean the stickers adhered better to the jack Some
trimming was still required of each sticker to avoid them adding extra
resistance to the jack's movement.

During testing it was found that their was a considerable amount
cross-talk between adjacent sensors. This meant the reading from key
would be affected by the movement of the key next to it. The cross-talk
was assumed to be as a result of light reflecting from other jacks.
Given that IR is used this was difficult to confirm by sight. A set of
baffles was designed for the PCB that would wrap round the jack and stop
light from other jacks or sensors. The baffles were printed on a BAMBU X1 3D
printers using PLA plastic. A dark pigmented PLA was used to avoid IR
light simply being diffused. After the baffles were fitted, the
cross-talk was eliminated.

\subsubsection{RGB LEDs}\label{rgb-leds}

The LEDs also allowed for easy identification for keys that were outside
of expected values and when a key was triggered. This allowed for visual
identification of malfunctioning or un-calibrated keys.

The LEDs are on the same PCB as the sensors and are hidden when the
name board is in place.2024

Would suggest that this is a necessity of this kind to provide a means
to identify and diagnose LEDs with state.

TLC5940 or other PWM driver such as in \cite{McPherson2013} added
complexity to assembling, but were avoid in favour of Addressable LEDs
with integrated driver. The integrated LEDs share the same data line
which has the limitation where a faulty LED will stop all following LEDs
from working. An individual PWM driver per PCB would allow avoid
situations where failure cascades, but would add complexity in assembly
as the parts are typically SMD. Also, given the LEDs are fixed to the
board and could not be repaired in situ, the solution would be to
replace the PCB regardless of which approach was taken. Given the LEDs
were only used in debugging and calibration and otherwise would not be
visible, a simpler design approach was taken.

\subsubsection{Multiplexing}\label{multiplexing}

The BLE Nano has 8 ADC pins connected to one ADC channel via a multiplexer. 
We avoided the complexity of communicating multiple microcontrollers and instead used a Texas Instruments CD4051B multiplexer on each sensor board.
An 8-channel multiplexer was chosen as the model has 49-keys and a division of 7 PCBs each containing 7 sensors.
The multiplexer reduced the 7 sensor signals to a single channel. The CD4051B is addressed with a 3-bit which necessitates 3 digital channels.
Transistor / Op-amp circuit could have been used similar to the one in
\cite{McPherson2013}. Given time constraints it was beneficial to continue using and entirely THT based designed for easy assembly.

% \subsection{PCB design}\label{pcb-design}

% EAGLE PCBs designed using EAGLE.

% Worth noting that AutoCAD has
% deprecated EAGLE and the programme will no longer be supported from
% 2026 Project files will need to be transitioned to another standard,
% likely the KiCAD programme.

% trade off between limiting number of boards required and allowing
% compensation for variation in pitch.

% Smaller PCBs means that spacing could be adjusted to accomodate the
% changes in pitch between jacks. It also allowed for a modular approach
% where any PCB could swapped for minimal cost.

% \begin{itemize}
% \item
%   x7 QRE1113
% \item
%   x7 100R
% \item
%   x7 10kR
% \item
%   x1 CD4051BE Multipexer
% \item
%   jack / sensor pitch by board
% \item
%   7 boards adjust for pitch/spacing error
% \item
%   modular, replaceable
% \item
%   orientation problems

%   \begin{itemize}
%   \item
%     legs break easily
%   \item
%     end mount like A.P.Mc. time consuming
%   \item
%     problem

%     \begin{itemize}
%     \item
%       milling out keys
%     \end{itemize}
%   \end{itemize}
% \item
%   designed from distance
% \item
%   striped lines across pcbs
% \item
%   problems

%   \begin{itemize}
%   \item
%     additional terminals uneeded resulted in lots of additional work
%   \item
%     long cabling increase risk of breaking
%   \end{itemize}
% \item
%   solutions

%   \begin{itemize}
%   \item
%     striping all ADC lines with FFC
%   \item
%     solder pads to select channel
%   \end{itemize}
% \end{itemize}

@@

Firmware checks components are connected then reads the FRAM to see if
threshold data is already present. The rotary encoder and it's momentary
switch are polled for any change in state. A key is selected with the
rotary encoder while the momentary switch is used for changing between
states for key selection and threshold adjustment A double click of the
rotary encoder writes the current threshold values to FRAM. The sensors
for the current multiplexer channels are polled. If any sensor reads
above the threshold when it was previously below a MIDI On message is
sent. For the reverse case where the value was below the threshold when
it was previously above, a MIDI Note Off message is sent.

@@

The final assembly took place at the NEMUS lab at the University of
Bologna over the final week in October 2024 for delivery at San
Colombano for exhibition in the first week of November in 2024.

@@

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{src/images/keys-milled.jpeg} 
  \caption{} 
  \Description{} 
  \label{fig:milling}
\end{figure}

@@

The PCBs shared power and data lines for components with the exception
of the output from the multiplexer. Ribbon cable was cut to size to
allow for some room to readjust the distance of the sensors from the
jacks. The data signals from each PCB were connected using a separate
cable loom. Soldering cables and inspecting each join required another
day work. This did allow for readjustments to the position of the
controller board Measurements of the rear chamber did not match with the
Mac Mini that was to be used a different layout was decided for fitting
the controller board/ Cutting cables to size allowed for a great deal of
flexibility but at the cost of time. After the first install it was
clearer what constraints were in place for fitting and an alternative
wiring system using flat flex cable was implement (discussed in
\textbackslash section)

@@

As the PCBs were fitted the jacks had the gradient stickers applied to
them. The jacks had been coated with a varnish on the short side to
provide a better surface for the stickers to adhere to. After the
stickers were applied excess material was trimmed from each jack. The
jack were then check that they moved freely in their slot.

After fitting the PCBs preliminary calibration was carried out. Before
the keys were refitted the jacks were put back in place. The data from
each key was confirmed to be in a similar range. Any key showing
anomalous data had it's jack removed and re-inspected. The PCB fitting
process and refitting required another day as faults were found in
wiring or misalignment of the stickers. After the sensors were
confirmed to be functioning the keys were put back.

The calibration workflow was with \anon{Dr. Craig Webb} who was unfamiliar with
the system. Adjustments were made to the firmware to to make it more
intuitive for the user. Addressable LEDs became vital as it allowed for
easy identification of the key being calibrated. They were also used to
identify any keys whose current reading was beyond what was expected.

@@

A process was implemented where the key was selected by playing, but
this caused problems in situations where the data was unusual due to
misalignment of the sensors.

This would likely be a better system for when the keyboard was fully
operational.

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{images/serial_monitor.png} 
  \caption{} 
  \Description{} 
  \label{fig:serial_monitor}
\end{figure}

The thresholds were calibrated visually utilising the serial plotter of
the Arduino IDE (Figure \ref{fig:serial_monitor}) and the open source MIDI
Monitor software \footnote{\url{https://github.com/krevis/MIDIApps}}.

@@

Finer calibration carried out with the guidance of \anon{Catalina Vincens} as
an expert harpsichord performer

@@

A limitation in the design is the variability in positioning of tag,
especially when they need to be reapplied. A jig was designed for
applying tags, but could not be printed in time for installation. Even
with a jig, some tags had to be readjusted by hand regardless. Long term
this has implications on how to maintain the interface.

Taken a different approach to what surface is used. Preliminary test
were taken using Laser Cutting and CNC to fabricate Jack bodies. Laser
cutting results in a loss of material that has too much variability. The
edge of the jack body will be charred. The material also needs to be
reliable. on the advice of two luthiers, \anon{Roberto Livi} and \anon{Jonathan Santa
Maria Bouquet} at \anon{Edinburgh}, Plywood and MDF have problems with
variations in humidity, tear-out and delamination.

Some tests were carried out using a notched jack body that was 3D
printed. The notch means that as the jack raises, the forward material
acts like a shutter. No reasons that this could not be CNC though a
variety of materials would be needed. Traditional material used is xx
though need to test of the material is naturally reflective enough or if
a coating would need to be applied.

@@

A second interface designed more closely to the Trasuntino style
instrument commissioned by the \anon{NEMUS} project with \anon{Roberto Livi}. Using
lessons learned for the first iteration the second can make better
accommodations for internal electronics. Interface to be used for a
performer study and used for performance with a numerically simulated
sound synthesis engine.

@@

Future novel designs for an interface that would utilise velocity and
aftertouch messages in software. This would be applicable for numerical
simulations of strings. A bespoke piece of software would also allow
for direct mapping of jack rows to registers. There is enough data to
implement functionality for velocity and aftertouch MIDI messages and
simply needs to be activated in the firmware. (Hyperinstruments
\cite{nime2024_20})

@@

Using the Espressif development framework ESP-IDF and following the same
logic as the original firmware source, the latency was reduced to around
10ms. ESP-IDF allows access to functionality of the ESP32 that is not
exposed in the Arduino libraries. It is possible that the 10ms latency
could be improved upon, but there would be a cost in the form of
development time. Consideration would need to be made on how to port
existing code or whether to completely re-write the firmware.

@@

  \begin{itemize}
  \item
    debug if problem is power related Add power small power LEDs to the
    each PCB. It was not common for a board to not be functioning and
    thye result to be that the power was not correctly connected.
    Especially when the power is connected via something such as an FFC
    cable where it is easy to plug-in the wrong way especially when
    fitting is to be carried out by someone unfamiliar with the system.
    Power LEDs would quickly confirm whether the problem was power
    related. They would have to be small enough and low enough current
    draw to avoid glare or being mistaken with other indicator LEDs To
    continue using THT components, the choice would have to be made
    carefully.
  \end{itemize}
\end{itemize}
